---
title: "ar1"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

## AR(1) models in `inlabru`

In this exercise we will:

-   Fit an AR(1) process with `inlabru` to model lakes water levels

-   Change the default priors for the observational error

-   Set penalized complexity priors for the correlation and precision parameters of the latent effects.

-   Fit a RW(1) model

Start by loading useful libraries:

```{r}
library(tidyverse) 
library(INLA) 
library(ggplot2)
library(patchwork) 
library(inlabru)
library(DAAG)
```

## Great Lakes water level

In this exercise we will look at [greatLakes](https://rdrr.io/cran/DAAG/man/greatLakes.html) dataset from the `DAAG` package. The data set contains the water level heights for the lakes Erie, Michigan/Huron, Ontario and St Clair from 1918 to 2009.

Lets begin by loading and formatting the data into a tidy format.

```{r}
data("greatLakes")

greatLakes.df = data.frame(as.matrix(greatLakes),
                           year = time(greatLakes)) %>%
  pivot_longer(cols = c("Erie","michHuron","Ontario","StClair"),
               names_to = "Lakes",
               values_to = "height" ) 

```

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 6
#| fig-height: 5


greatLakes.df %>%
  ggplot(aes(x=year,y=height)) + 
  geom_point() +
  facet_wrap(~Lakes,scales="free")

```

## Fitting an AR(1) model in `inlabru`

We will focus on the Erin lake for now. Lets begin by fitting an AR(1) model of the form:

$$
\begin{aligned}
\text{height}_t &= \alpha + u_t +\varepsilon_t~; ~~ \varepsilon_t\sim \mathcal{N}(0,\tau_e^{-1}) \\
u_t &= \phi u_{t-1} + \delta_t~~~ ; ~~ \delta_t \sim \mathcal{N}(0,\tau_u^{-1}); ~~ t > 1 \\
x_1 &= \mathcal{N}(0,\kappa^{-1})\\
\kappa &= \tau_u (1-\phi^2)
\end{aligned}
$$

Where $\alpha$ is the intercept, $\phi$ is the correlation term, $\varepsilon$ is the observational Gaussian error with mean zero and precision $\tau_e$ and $\kappa$ is the marginal precision for the state $u_t$ for $t= 1,\ldots,92$.

First we make a subset of the dataset and create a time index $T$:

```{r}
greatLakes.df$t.idx <- greatLakes.df$year-1917

Erie.df = greatLakes.df %>% filter(Lakes == "Erie")

```

Now we fit model in `inlabru` by defining (1) the model components (2) the formula and (3) the observational model.

```{r}
# Model components
cmp =  ~ -1 + alpha(1) + ut(t.idx,model = "ar1")
# Model formula
formula = height ~ alpha + ut


# Observational model
lik =  bru_obs(formula = height   ~.,
            family = "gaussian",
            data = Erie.df )

# fit the model
fit.Erie_ar1 = bru(cmp, lik)

```

**Model predictions**

```{r}

pred_ar1.Erie = predict(fit.Erie_ar1, Erie.df, ~ alpha + ut)

ggplot(pred_ar1.Erie,aes(y=mean,x=year))+
  geom_line()+
    geom_ribbon(aes(x = year, y = mean, ymin = q0.025, ymax = q0.975),
                alpha = 0.5) +
  geom_point(aes(y=height,x=year))
```

It is clear that the model overfits the data, leading to poor predictive performance. Thus, we need to introduce some prior information on the what we expect the variation of the process to be. Let review INLA' prior parametrization for autoregressive models:

**Priors**

Let $\pmb{\theta} = \{\theta_y,\theta_u,\theta_\phi\}$ be INLA's internal representation of the hyperparameters such that:

$\theta_y = \log(\tau^2_y)$

$\theta_u = \log(\kappa) = \log\left(\tau_u[1-\phi^2]\right)$

$\theta_\phi = \log \left(\frac{1+\phi}{1-\phi}\right)$

The default priors for $\{\theta_y,\theta_u\}$ are $\text{log-gamma} (1, 5\times 10^{-5} )$ priors with default initial values set to 4 in each case. Then, Gaussian priors $\alpha \sim \mathcal{N}(0,\tau_y = 0.001)$ and $\theta_\phi \sim \mathcal{N}(0, \tau_y= 0.15)$ are used for the intercept and correlation parameter respectively.

::: callout-note
Specifically for AR(1) correlation parameter $\phi$, INLA uses the following logit transformation on $\theta_\phi$:

$$ \phi = \frac{2\exp(\theta_\phi)}{1+ \exp(\theta_\phi)} -1. $$
:::

**Setting priors and PC-priors**

Lets now set a Gamma prior with parameters 1 and 1, so that the precision of the Gaussian osbervational error is centered at 1 with a variance of 1. Additionally we will set Penalized Complexity (PC) priors according to the following probability statements:

-   $P(\sigma > 1) = 0.01$

-   $P(\phi > 0.5) = 0.3$

Notice that the PC prior for the precision $\tau_u$ is defined on the standard deviation $\sigma_u = \tau_u^{-1/2}$

```{r}
pc_prior <- list(theta = list(prior = "pc.prec", param = c(1, 0.01)),
                 rho = list(prior = "pc.cor0", param = c(0.5, 0.3))) 

prec.tau_e <- list(prec = list(prior = "loggamma",   # prior name
                             param = c(1, 1))) # prior values

# Model components
cmp =  ~ -1 + alpha(1) + ut(t.idx, model = "ar1",  hyper = pc_prior)
# Model formula
formula = height ~ alpha + ut


# Observational model
lik =  bru_obs(formula = height  ~.,
            family = "gaussian",
            data = Erie.df,
            control.family = list(hyper = prec.tau_e))

# fit the model
fit.Erie_ar1 = bru(cmp, lik)

# Model predictions
pred_ar1.Erie = predict(fit.Erie_ar1, Erie.df, ~ alpha + ut)



```

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 4
#| fig-height: 4
#| 
ggplot(pred_ar1.Erie,aes(y=mean,x=year))+
  geom_line()+
    geom_ribbon(aes(x = year, y = mean, ymin = q0.025, ymax = q0.975),
                alpha = 0.5) +
  geom_point(aes(y=height,x=year))
```

## Fitting a RW(1) model

Now we fit a random walk of order 1 to the Erie lake data:

$$
\begin{aligned}
y_t &= \alpha + u_t + \varepsilon_t, ~ t = 1,\ldots,92 \\
 \varepsilon_t & \sim \mathcal{N}(0,\tau_e) \\
u_t - u_{t-1} &\sim \mathcal{N}(0,\tau_u),~ t = 2,\ldots,92 \\
\end{aligned}
$$

Firs we define model priors:

```{r}

pc_prior <- list(theta = list(prior = "pc.prec", param = c(1, 0.01))) 

prec.tau_e <- list(prec = list(prior = "loggamma",   # prior name
                             param = c(1, 1))) # prior values

```

Now we define model components:

```{r}
cmp_rw =  ~ -1 + alpha(1) + 
  ut(t.idx ,
     model = "rw1",
     constr = FALSE,
     hyper=pc_prior,
     scale.model = TRUE)
```

Notice that we have set `scale.model = TRUE` to scale the latent effects. This is particularly important when Intrinsic Gaussian Markov random fields (IGMRFs) are used as priors (e.g., random walk models or some spatial models) for the latent effects. By defining `scale.model = TRUE`, the `rw1`-model is scaled to have a generalized variance equal to one. By scaling scaling the models we ensure that a fixed hyperprior for the precision parameter has a similar interpretation for different types of IGMRFs, making precision estimates comparable between different models. Scaling also allows estimates to be less sensitive to re-scaling covariates in the linear predictor and makes the precision invariant to changes in the shape and size of the latent effect (see @sørbye2014 for further details) .

We can now fit the model with the updated components and plot the predicted values

```{r}
# fit the model
fit.Erie_rw1 = bru(cmp_rw, lik)
# Model predictions
pred_rw1.Erie = predict(fit.Erie_rw1, Erie.df, ~ alpha + ut)

```

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 4
#| fig-height: 4

ggplot(pred_rw1.Erie,aes(y=mean,x=year))+
  geom_line()+
    geom_ribbon(aes(x = year, y = mean, ymin = q0.025, ymax = q0.975),
                alpha = 0.5) +
  geom_point(aes(y=height,x=year))

```

## Group-level effects

Now we will model the height water levels for all four lakes by grouping the random effects. This will allow a within-lakes correlation to be included. In the next example, we allow for correlated effects using an `ar1` model for the years and `iid` random effects on the lakes. First we create a lakes id and set the priors for our model:

```{r}
greatLakes.df$lake_id <- as.numeric(as.factor(greatLakes.df$Lakes))

pc_prior <- list(theta = list(prior = "pc.prec", param = c(1, 0.01)),
                 rho = list(prior = "pc.cor0", param = c(0.5, 0.3))) 

prec.tau_e <- list(prec = list(prior = "loggamma",   # prior name
                             param = c(1, 10))) # prior values

```

Now we define the model components. The lakes IDs that define the group are passed with parameter `group` argument and the `iid` model and other parameters are passed through the `control.group` parameter.

```{r}
# Model components
cmp =  ~ -1 + alpha(1) + ut(year,model = "ar1",
                            hyper = pc_prior,
                            group =lake_id,
                            control.group = 
                            list(model = "iid", 
                                 scale.model = TRUE))
```

We fit the model in a similar fashion as we did before:

```{r}
# Model formula
formula = height ~ alpha + ut


# Observational model
lik =  bru_obs(formula = height  ~.,
            family = "gaussian",
            data = greatLakes.df,
            control.family = list(hyper = prec.tau_e))

# fit the model
fit.all_lakes_ar1 = bru(cmp, lik)

# Model predictions
pred_ar1.all = predict(fit.all_lakes_ar1, greatLakes.df, ~ alpha + ut)

```

Lastly we can visualize group-level model predictions as follows:

```{r}
ggplot(pred_ar1.all,aes(y=mean,x=year))+
  geom_line()+
    geom_ribbon(aes(x = year, y = mean, ymin = q0.025, ymax = q0.975),
                alpha = 0.5) +
  geom_point(aes(y=height,x=year)) + facet_wrap(~Lakes,scales = "free")

```

## Non-Gaussian data

In the next example we will use the `Toyo` data set to illustrate how temporal models can be fit to non-Gaussian data. The `Tokyo` data set available in `INLA` contains the recorded days of rain above 1 mm in Tokyo for 2 years, 1983:84. The data set contains the following variables:

-   `y` : number of days with rain

-   `n` : total number of days

-   `time` : day of the year

```{r}
data("Tokyo")
```

A possible observational model for these data is

$$
\begin{aligned}
y_t|\eta_t & \sim\text{Bin}(n_t, p_t) \\
\eta_t &= \text{logit}(p_t),\qquad i = 1,\dots,366
\end{aligned}
$$ $$
n_t = \left\{
 \begin{array}{lr}
1, & \text{for}\; 29\; \text{February}\\
2, & \text{other days}
\end{array}\right.
$$ $$
y_t =
\begin{cases}
\{0,1\}, & \text{for}\; 29\; \text{February}\\
\{0,1,2\}, & \text{other days}
 \end{cases}
$$

Then, the latent field is given by

$$
\eta_t = \beta_0 + f(\text{time}_t)
$$

-   Where the probability of rain depends on on the day of the year $t$

-   $\beta_0$ is an intercept

-   $f(\text{time}_t)$ is a temporal model, e.g., a RW2 model (this is just a smoother).

The smoothness is controlled by a hyperparameter $\tau_f$ . Thus, we assign a prior to $\tau_f$ to finalize the model.

We can fit the model as follows:

```{r}

# define model component
cmp =  ~ -1 + beta0(1) + time_effect(time, model = "rw2", cyclic = TRUE)

# define model predictor
eta = y ~ beta0 + time_effect

# build the observation model
lik = bru_obs(formula = eta,
              family = "binomial",
              Ntrials = n,
              data = Tokyo)

# fit the model
fit = bru(cmp, lik)
```

Notice that we have set `cyclic = TRUE` as this is a cyclic effect. Finaly, we can produce model predictions in a similar fashion as we did before:

```{r}

pTokyo = predict(fit, Tokyo, ~ plogis(beta0 + time_effect))

ggplot(data=pTokyo , aes(x= time, y= y) ) +
  geom_point() + 
  ylab("") + xlab("") +
  # Custom the Y scales:
  scale_y_continuous(
    # Features of the first axis
    name = "",
    # Add a second axis and specify its features
    sec.axis = sec_axis( transform=~./2, name="Probability")
  )  + geom_line(aes(y=mean*2,x=time)) +
  geom_ribbon(aes( ymin = q0.025*2, 
                             ymax = q0.975*2), alpha = 0.5)
  
```
