---
title: "Spatial Data types"
format: 
  html:
    theme:
      light: flatly
      dark: darkly
  PrettyPDF-pdf:
    keep-tex: true
    number-sections: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

## Spatial Data {#sec-spat_data}

```{r}
#| echo: false
#| warning: false
#| message: false
#| purl: false

library(webexercises)

```

In this practical we will:

-   specific ILO1
-   specific ILO2

Start by loading useful libraries:

```{r}
#| warning: false
#| message: false
#| code-summary: "Load libraries"

library(dplyr)
library(INLA)
library(ggplot2)
library(patchwork)
library(inlabru)     
# load some libraries to generate nice plots
library(scico)
```

As our first example we consider a simple linear regression model with Gaussian observations

$$
y_i\sim\mathcal{N}(\mu_i, \sigma^2), \qquad i = 1,\dots,N
$$

::: {.callout-tip icon="false"}
## {{< bi question-octagon color=#6dc83c >}} Question

What are the three spatial data type?

`r hide("Answer")`

Geostatistical data, Areal data and Point processes .

`r unhide()`
:::

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Task description

`r hide("hint")`

This is a hint

`r unhide()`

```{r}
#| webex.hide: "Click here to see the solution"
#| purl: false

2+2

```
:::

```{r}
#| echo: false
#| eval: false

iris_sub<- iris[1:5,1:2]
write.csv(iris_sub,file = "iris_subset.csv",row.names = F)


library(downloadthis)

iris_sub %>% download_this(output_name = "iris subset data",
                           output_extension = ".csv",
                           button_label = "Download data",
                           button_type = "success",
                           has_icon = TRUE,
                           icon = "fa fa-save")

```


{{< downloadthis iris_subset.csv dname = "iris subset data" label ="Download data" icon="database-fill-down" type = "success">}}
