---
title: "BES workshop"
author: "Jafet Belmont, Sara Martino and Janine Illian"
subtitle: "British Ecological Society Meeting 2025, Edinburgh"
---

# Welcome to the course!

-   Welcome to the `inlabru` workshop!
-   The goal of this workshop is to introduce new users to the `inlabru` software for Bayesian spatial modelling using the integrated nested Laplace approximation (INLA).
-   The workshop is intended for applied statisticians and quantitative ecologists who are interested in efficiently modelling spatial data- No prior knowledge of `R-INLA` is required.

# Learning Objectives for the workshop

At the end of the workshop, participants will be able to:

-   Understand the `inlabru` workflow and software principles.
-   Fit commonly used models such as Linear models, LMM, GLM and GAMs using `inlabru`.
-   Describe different types of spatial data and fit appropriate spatial models.
-   Interpret model outputs and create spatial maps and predictions.
-   Implement a complete, reproducible spatial analysis workflow from data preparation to result visualization.

# Schedule Program

| Time | Topic |
|:----------------------------------:|:----------------------------------:|
| 13:00 - 14:00 | Workshop registration |
| 14:00 - 14:45 | [**Session 1**](https://ecol-stats.github.io/inlabru_workshop/BES_workshop/slides/slides_2#/title-slide): Introduction to `inlabru` |
| 14:45 - 15:45 | [Practical Session 1](day1_practical.qmd) |
| 15:45 - 16:00 | Coffee break |
| 16:00 - 16:45 | [**Session 2**:](https://ecol-stats.github.io/inlabru_workshop/BES_workshop/docs/slides/slides_2) Spatial Modelling with `inlabru` |
| 16:45 - 17:45 | [Practical Session 2](practical2_compiler.qmd) |
| 17:45 - 18:00 | Wrap-up and outlook |

: {.striped .hover}

# In preparation for the workshop

Participants are required to follow the next steps before the day of the workshop:

1.  Install [R-INLA](https://www.r-inla.org/download-install)

2.  Install `inlabru` (available from CRAN)

    ```{r}
    #| eval: false

    # Enable universe(s) by inlabru-org
    options(repos = c(
      inlabruorg = "https://inlabru-org.r-universe.dev",
      INLA = "https://inla.r-inla-download.org/R/testing",
      CRAN = "https://cloud.r-project.org"
    ))

    # Install some packages
    install.packages("inlabru")
    ```

3.  Make sure you have the latest `R-INLA`, `inlabru` and `R` versions installed.

4.  Install the following libraries

    ```{r}
    #| echo: false
    #| warning: false
    #| message: false

    library(DT)  # For interactive tables

    if (file.exists("practicals_libraries.rds")) {
      libs_data <- readRDS("practicals_libraries.rds")
      
      if (nrow(libs_data) > 0) {
        cat("### Libraries Used in Practicals\n\n")
        cat("We use **", nrow(libs_data), "** unique libraries across all exercises.\n\n")
        
        # Create interactive table
        datatable(
          libs_data,
          options = list(
            pageLength = 10,
            dom = 'tip',
            autoWidth = TRUE
          ),
          colnames = c('Library', 'Used In Files'),
          caption = 'Libraries used in practical exercises'
        )
      }
    }

    ```
