---
title: "`inlabru` Workshop"
author: "Jafet Belmont, Sara Martino and Janine Illian"
subtitle: "ICES 2025, Copenhagen"
format: 
  html: default
  PrettyPDF-pdf:
    keep-tex:  true
    number-sections: true
editor_options: 
  chunk_output_type: console
---

# Welcome to the course!

-   Welcome to the `inlabru` workshop!
-   The goal for the workshop is to ...
-   The workshop is intended for ... No knowledge of `R-INLA` is required.
-   Workshop materials in the github repository [inlabru-workshop](https://github.com/Ecol-Stats/inlabru_workshop/tree/main/docs/copenague)

# Learning Objectives for the workshop

At the end of the workshop, participants will be able to:

-   ILO1
-   ILO2
-   ILO3

Intended audience and level: The tutorial is intended for ... No knowledge of `R-INLA` is required.

# Schedule Program {#schedule-tabs}

::: panel-tabset
### Day 1

|     Time      |                     Topic                      |
|:-------------:|:----------------------------------------------:|
| 10:00 - 10:30 |            ICES informative session            |
| 10:30 - 11:30 |    **Session 1**: Introduction to `inlabru`    |
| 11:30 - 13:00 |   [Practical Session 1](day1_practical.qmd)    |
| 13:00 - 14:30 |                 Lunch break 🍽️                 |
| 14:30 - 15:30 | **Session 2**: Latent Gaussian Models and INLA |
| 15:30 - 15:45 |                Coffee Break ☕                 |
| 15:45 - 16:45 |  [Practical Session 2](day1_practical_2.qmd)   |
| 16:45 - 17:00 |              wrap-up and outlook               |

: {.striped .hover}

### Day 2

|     Time      |                         Topic                          |
|:-------------:|:------------------------------------------------------:|
| 9:00 - 10:00  | **Session 3**: Temporal modelling and smoothing part 1 |
| 10:00 - 10:30 |                        Snack 🥙                        |
| 10:30 - 11:30 | **Session 4**: Temporal modelling and smoothing part 2 |
| 11:30 - 13:00 |      [Practical Session 3](day2_practical_3.qmd)       |
| 13:00 - 14:30 |                     Lunch break 🍽️                     |
| 14:30 - 15:30 |   **Session 5**: Introduction to Spatial Statistics    |
| 15:35 - 15:45 |                    Coffee Break ☕                     |
| 15:45 - 16:45 |      [Practical Session 4](day2_practical_4.qmd)       |
| 16:45 - 17:00 |                  wrap-up and outlook                   |

: {.striped .hover}

### Day 3

|     Time      |                         Topic                         |
|:-------------:|:-----------------------------------------------------:|
| 9:00 - 10:00  |            **Session 6**: Areal Processes             |
| 10:00 - 10:30 |                       Snack 🥙                        |
| 10:30 - 11:30 |             **Session 7**: Geostatistics              |
| 11:30 - 13:00 |      [Practical Session 5](day3_practical_5.qmd)      |
| 13:00 - 14:30 |                    Lunch break 🍽️                     |
| 14:30 - 15:30 |        **Session 8**: Spatial Point processes         |
| 15:35 - 15:45 |                    Coffee Break ☕                    |
| 15:45 - 16:45 | [Practical Session 5 continued](day3_practical_5.qmd) |
| 16:45 - 17:00 |                  wrap-up and outlook                  |

: {.striped .hover}

### Day 4

|     Time      |                       Topic                       |
|:-------------:|:-------------------------------------------------:|
| 9:00 - 10:00  |       **Session 9**: Spatiotemporal models        |
| 10:00 - 10:30 |                     Snack 🥙                      |
| 10:30 - 11:30 |  **Session 10**: Model comparison and evaluation  |
| 11:30 - 13:00 |   [Practical Session 6](day4_practical_6.qmd)     |
| 13:00 - 14:30 |                  Lunch break 🍽️                   |
| 14:30 - 15:30 | **Session 11**: Multilikelihoods/joint likelihood |
| 15:35 - 15:45 |                  Coffee Break ☕                  |
| 15:45 - 16:45 |                Practical Session 7                |
| 16:45 - 17:00 |                wrap-up and outlook                |

: {.striped .hover}

### Day 5

| Time | Topic |
|:----------------------------------:|:----------------------------------:|
| 9:00 - 10:00 | **Session 12**: Zero inflated models |
| 10:00 - 10:30 | Snack 🥙 |
| 10:30 - 11:30 | **Session 13**: Complex observational processes: Distance Sampling |
| 11:30 - 13:00 | Practical Session 8 |
| 13:00 - 13:15 | Coffee Break ☕ |
| 13:15 - 14:00 | Closing session |

: {.striped .hover}
:::

\clearpage

# In preparation for the workshop

Participants are required to follow the next steps before the day of the workshop:

1.  Install [R-INLA](https://www.r-inla.org/download-install)

2.  Install `inlabru` (available from CRAN)

```{r}
#| eval: false
# Enable universe(s) by inlabru-org
options(repos = c(
  inlabruorg = "https://inlabru-org.r-universe.dev",
  INLA = "https://inla.r-inla-download.org/R/testing",
  CRAN = "https://cloud.r-project.org"
))

# Install some packages
install.packages("inlabru")
```

3.  Make sure you have the latest `R-INLA`, `inlabru` and `R` versions installed.
4.  Install the following libraries:

```{r}
#| results: hide
#| echo: false

source("extract_practicals_libraries.R")

```

```{r}
#| echo: false
#| warning: false
#| message: false


    if (file.exists("practicals_libraries.rds")) {
      libs_data <- readRDS("practicals_libraries.rds")
      
      # Libraries to exclude
      exclude_libs <- c("webexercises", "INLA", "inlabru", "knitr")
      
      # Filter out excluded libraries
      filtered_libs <- libs_data[!libs_data$Library %in% exclude_libs, ]
      
      if (nrow(filtered_libs) > 0) {

        cat("install.packages(c(\n")
        cat(paste0('  "', filtered_libs$Library, '"', collapse = ",\n"))
        cat("\n))")
        
      } else {
        cat("No libraries found after filtering.\n")
      }
    }
```
